Why Real-World Subgraph Isomorphism is Easy, \\ and What This Means for Graph Databases

### 2016-09-21 paper/paper.tex
Make a start on this

\documentclass[twoside,11pt]{article}
\usepackage{jair}
\usepackage{theapa}
\usepackage{rawfonts}
\usepackage{complexity}
\usepackage{cleveref}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{amsmath}
\usepackage{amssymb}

@@

\newcommand{\citet}[1]{\citeauthor{#1} \shortcite{#1}}
\newcommand{\citep}[1]{\cite{#1}}

@@

\title{Why Real-World Subgraph Isomorphism is Easy, \\ and What This Means for Graph Databases}

@@

    When labels are introduced, as commonly seen in graph databases, additional rich behaviour
    arises.  We show that when only slightly better algorithms are used, all commonly used benchmark
    suites for evaluating graph database systems become trivial, and that existing notions of
    difficulty are based almost entirely upon unnecessarily poor performance from a widely used
    subgraph isomorphism algorithm. Finally, we argue that popular indexing techniques used in graph
    databases are flawed, and would not give any benefits even with harder instances when used with
    any reasonable modern subgraph isomorphism algorithm.
\end{abstract}

\section{Introduction}

The \emph{non-induced subgraph isomorphism problem} is to find an injective mapping from a given
pattern graph to a given target graph which preserves adjacency---in essence, we are ``finding a
copy of'' the pattern inside the target. The \emph{induced} variant of the problem additionally
requires that the mapping preserve non-adjacency, so there are no ``extra edges'' in the copy of the
pattern that we find. We illustrate both variants in \cref{figure:sip}.
Despite these problems being \NP-complete, modern practical subgraph isomorphism algorithms can
handle problem instances with many hundreds of vertices in the pattern graph, and up to ten thousand
vertices in the target graph \cite{Cordella:2004,Solnon:2010,Audemard:2014,McCreesh:2015}, leading
to successful application in areas such as computer
vision \cite{Damiand:2011,Solnon:2015}, biochemistry \cite{Giugno:2013,Carletti:2015}, and pattern
recognition \citep{Conte:2004}.

\begin{figure}[t]
    \centering
    \begin{tikzpicture}[scale=0.5,every node/.style={font=\footnotesize}]
        \begin{scope}
            \node [draw, circle, fill=uofgcobalt, inner sep=1.5pt] (Na) at (1,  0) {1};
            \node [draw, circle, fill=uofgpumpkin, inner sep=1.5pt] (Nb) at (1, -1.5) {2};
            \node [draw, circle, fill=uofglawn, inner sep=1.5pt] (Nc) at (0, -3) {3};
            \node [draw, circle, fill=uofgthistle, inner sep=1.5pt] (Nd) at (2, -3) {4};

@@

            \node [anchor=center, font=\Large] (A1) at (-1.5, -1.5) { $\hookleftarrow$ };
            \node [anchor=center, font=\Large] (A2) at ( 3.5, -1.5) { $\rightarrowtail$ };
        \end{scope}
    \end{tikzpicture}\\[0.5cm]
    \begin{tikzpicture}[scale=0.45,every node/.style={font=\footnotesize}]
        \begin{scope}
            \foreach \n in {1, ..., 6}{
                \node [draw, circle, fill=uofgcobalt, inner sep=1pt] (N1\n) at (\n, 0) {\n};
            }

@@

    \caption{On top, there is a non-induced isomorphism from the pattern graph, in the center, to the target
    on the right, mapping vertex 1 to 1, 2 to 3, 3 to 5 and 4 to 6. This is not an induced
    isomorphism, since there is an edge between 1 and 5 in the target but not between 1 and 3 in the
    pattern. The mapping from the pattern to the (same) target on the left, sending 1 to 2, 2 to 6, 3 to
    5 and 4 to 3, is both non-induced and induced.
    Below, the association graph encoding for the induced version: the highlighted clique
    corresponds to the same mapping.}
    \label{figure:sip}
\end{figure}

However, these algorithms cannot handle \emph{arbitrary} instances of this size. The experimental
evaluations of these algorithms were performed using a mix of real-world graphs, graphs that encode
biochemistry and computer vision problems, and randomly generated graph pairs. Using random
instances to evaluate algorithm behaviour can be beneficial, because it provides a way of generating
many instances cheaply, and reduces the risk of over-fitting when tuning design parameters. The
random instances used in each case came from common datasets \citep{DeSanto:2003,Zampelli:2010},
which were generated by taking a random subgraph of a random (Erd\H{o}s-R\'enyi, scale-free, bounded
degree, or mesh) graph and permuting the vertices. Such instances are guaranteed to be
satisfiable---\citet{Anton:2009} exploited this property to create large sets of random satisfiable
boolean satisfiability instances.  However, since this has been the only approach used to generate
random subgraph isomorphism instances, existing benchmark suites contain relatively few non-trivial
unsatisfiable instances, and the satisfiable instances tend to be computationally fairly easy, with
most of the difficulty being in dealing with the size of the model. This has lead to bias in
algorithm design, to the extent that some proposed techniques will \emph{only} work on satisfiable
instances \citep{Battiti:2007}.

The lack of unsatisfiable instances cannot be addressed by using a pattern graph from one of the
random suites with the ``wrong'' target graph: this tends to give either a trivially unsatisfiable
instance, or a satisfiable instance. (In particular, it is \emph{not} the case that a relatively
small random graph is unlikely to appear in a larger random graph.)

Here we present and evaluate a new method for creating random pattern/target pairs. This method
generates both satisfiable and unsatisfiable instances, and can produce computationally challenging
instances with only a few tens of vertices in the pattern, and 150 vertices in the target. Our work
builds upon the phase transition phenomena observed in satisfiability and graph colouring problems
first described by \citet{DBLP:conf/ijcai/CheesemanKT91} and \citet{Mitchell:1992}.  For subgraph isomorphism we
identify three relevant control parameters: we can independently alter the edge probability of the
pattern graph, the edge probability of the target graph, and the relative orders (number of
vertices) of the pattern and target graphs.  For non-induced isomorphisms, with the correct choice
of parameters we see results very similar to those observed with boolean satisfiability problems:
there is a phase transition (whose location we can predict) from satisfiable to unsatisfiable,
we see a solver-independent complexity peak occur near this phase transition, and understanding this
behaviour helps us to select variable and value ordering heuristics.

@@

@@

Throughout, our graphs are unlabelled, undirected, and do not have any loops.  The \emph{order} of a
graph is the cardinality of its vertex set. We write $\operatorname{V}(G)$ for the vertex set of a
graph $G$. The \emph{complement} of a graph $G$, denoted $\shortoverline{G}$, is the graph with the
same vertex set as $G$, and with an edge between distinct vertices $v$ and $w$ if and only if $v$
and $w$ are not adjacent in $G$. We write $G(n, p)$ for an Erd\H{o}s-R\'enyi random graph with $n$
vertices, and an edge between each distinct pair of vertices with independent probability $p$.

A \emph{non-induced subgraph isomorphism} from a graph $P$ (called the \emph{pattern}) to a graph
$T$ (called the \emph{target}) is an injective mapping from $\operatorname{V}(P)$ to
$\operatorname{V}(T)$ which preserves adjacency---that is, for every adjacent $v$ and $w$ in
$\operatorname{V}(P)$, the vertices $i(v)$ and $i(w)$ are adjacent in $T$. An \emph{induced subgraph
isomorphism} additionally preserves non-adjacency---that is, if $v$ and $w$ are not adjacent in $P$,
then $i(v)$ and $i(w)$ must not be adjacent in $T$. We use the notation $i : P \rightarrowtail T$
for a non-induced isomorphism, and $i : P \hookrightarrow T$ for an induced isomorphism. Observe
that an induced isomorphism $i : P \hookrightarrow T$ is a non-induced isomorphism $i : P
\rightarrowtail T$ which is also a non-induced isomorphism $\shortoverline{i} : \shortoverline{P}
\rightarrowtail \shortoverline{T}$.

@@

\begin{figure}[tb]
    %% \includegraphics{gen-graph-phase-transition.pdf}
    \caption{With a fixed pattern graph order of 20, a target graph order of 150, a target edge
        probability of 0.40, and varying pattern edge probability, we observe a phase transition and
        complexity peak with the Glasgow solver in the non-induced variant. Each point represents
        one instance. The lines show mean search effort and mean proportion satisfiable.}
    \label{figure:phase-transition}
\end{figure}

Suppose we arbitrarily decide upon a pattern graph order of 20, a target graph order of 150, and a
fixed target edge probability of 0.40. As we vary the pattern edge probability from 0 to 1, we would
expect to see a shift from entirely satisfiable instances (with no edges in the pattern, we can
always find a match) to entirely unsatisfiable instances (a maximum clique in this order and edge
probability of target graph will usually have between 9 and 12 vertices). The dark line in
\cref{figure:phase-transition} shows that this is the case. For densities of 0.67 or greater, no
instance is satisfiable; with densities of 0.44 or less, every instance is satisfiable; and with a
density of 0.55, roughly half the instances are satisfiable.

@@

\begin{figure}[tb]
    %% \includegraphics{gen-graph-non-induced.pdf}
    \caption{Behaviour of algorithms on the non-induced variant. For each plot, the x-axis is the
        pattern edge probability and the y-axis is the target edge probability, both from 0 to 1.
        Along the top row, we show the proportion of instances which are satisfiable; the white
        bands shows the phase transitions, and the black lines are our predictions of where the
        phase transition will occur. On the final three rows, we show the number of search nodes used by the
        Glasgow, LAD and VF2 solvers; the dark regions indicate ``really hard'' instances.}
    \label{figure:non-induced}
\end{figure}

@@

\begin{figure*}[p]
    %% \includegraphics{gen-graph-induced.pdf}
    \caption{Behaviour of algorithms on the induced variant, shown in the style
    of \cref{figure:non-induced}. The second, third and fourth rows show the number of search nodes used by the
    Glasgow, LAD and VF2 algorithms. The fifth row plots constrainedness: the darkest region is where
    $\kappa = 1$, and the lighter regions show where the problem is either over-
    or under-constrained. The final row shows when the Glasgow algorithm performs better when given
    the complements of the pattern and target graphs as inputs---the solid lines show the location
of the phase transition, and the dotted lines are $t_d=0.5$ and the $p_d=t_d$ diagonal.}\label{figure:induced}
\end{figure*}

\begin{figure*}[t]
    %% \includegraphics{gen-graph-sat.pdf}
    \caption{Behaviour of other solvers on the induced variant on smaller graphs, shown in the style of
        \cref{figure:non-induced}. The second row shows the number of search nodes used by the
    Glasgow algorithm, the third and fourth rows show the number of decisions made by the pseudo-boolean
and SAT solvers, and the final shows the number of search nodes used on the clique
encoding.}\label{figure:alt}
\end{figure*}

@@

For larger patterns, more complex behaviour emerges. If we are in the intersection of the bottom half
and the bottom right diagonal of the search space, we should always retain the original heuristic,
and if we are in the intersection of the top half and the top left diagonal, we should always use
the complements. This behaviour can be predicted by taking the partial derivatives of $\langle Sol
\rangle$ in the $-p_d$ and $t_d$ directions.  However, when inside the remaining two eighths of the
parameter space, the partial derivatives of $\langle Sol \rangle$ disagree on which heuristic to
use, and using directional derivatives is not enough to resolve the problem. A close observation of
the data suggests that the actual location of the phase transition may be involved (and perhaps
\citeauthor{Walsh:1998}'s suggestion applies only in these conditions). In any case, $\langle Sol
\rangle$ is insufficient to explain the observed behaviour in these two eighths of the parameter space.

@@

\section{Labelled Graphs}

\section{Other Models of Randomness}

\section{Techniques Used in Graph Databases}

>>>>

\documentclass[twoside,11pt]{article}
\usepackage{jair}
\usepackage{theapa}
\usepackage{rawfonts}
\usepackage{complexity}
\usepackage{cleveref}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{nicefrac}
\usepackage{csquotes}

% \usepackage{showframe}

@@

\newcommand{\citet}[1]{\citeA{#1}}
\newcommand{\citep}[1]{\cite{#1}}

@@

\title{Why Real-World Subgraph Matching is Easy, and \\ What This Means for Filter / Verify in
Graph Databases \thanks{Part of this paper appeared at IJCAI 2016
\citep{DBLP:conf/ijcai/McCreeshPT16}; \cref{section:introduction} has been expanded and
\cref{section:models,section:labelled,section:filterverify} are new material.}}

@@

    When labels are introduced, as commonly seen in graph databases, additional rich behaviour
    arises.  We show that when only slightly better algorithms are used, all commonly used benchmark
    suites for evaluating graph database systems become trivial, and that existing notions of
    difficulty are based almost entirely upon unnecessarily poor performance from a widely used
    subgraph isomorphism algorithm. With this in mind, we argue that widely researched ``filter /
    verify'' indexing technique used in graph databases is a flawed design based upon a widely
    repeated misunderstanding of what makes subgraph isomorphism hard. We explain why this technique
    would not be beneficial, even with harder instances than those currently used, when paired with
    any reasonable modern subgraph isomorphism algorithm.
\end{abstract}

\section{Introduction}\label{section:introduction}

The \emph{non-induced subgraph isomorphism problem} is to find an injective mapping from a given
pattern graph to a given target graph which preserves adjacency---in essence, we are ``finding a
copy of'' the pattern inside the target. The \emph{induced} variant of the problem additionally
requires that the mapping preserve non-adjacency, so there are no ``extra edges'' in the copy of the
pattern that we find. We illustrate both variants in \cref{figure:sip}.  Although these problems
are \NP-complete, modern practical subgraph isomorphism algorithms can handle problem instances
with many hundreds of vertices in the pattern graph, and up to ten thousand vertices in the target
graph \cite{Cordella:2004,Solnon:2010,Audemard:2014,McCreesh:2015}, leading to successful
application in areas such as computer vision \cite{Damiand:2011,Solnon:2015}, biochemistry
\cite{Giugno:2013,Carletti:2015}, and pattern recognition \citep{Conte:2004}.  In particular,
subgraph isomorphism is one of the key components in supporting complex queries in graph
databases---typically, these systems store a fixed set of target graphs, and for a sequence of
pattern queries, they must return every target graph which contains that pattern.  One common
approach to this problem is to combine a subgraph isomorphism algorithm with an indexing system in a
so-called ``filter / verify'' model, where invariants are used to attempt to pre-exclude
unsatisfiable instances to avoid the cost of a subgraph isomorphism call. In this context, the terms
\emph{matching} and \emph{verification} are often used for the subgraph isomorphism step (or a
slightly broader problem, for example permitting wildcards).

\begin{figure}[t]
    \centering
    \begin{tikzpicture}[scale=0.5,every node/.style={font=\footnotesize}]
        \begin{scope}[yshift=-2cm]
            \node [draw, circle, fill=uofgcobalt, inner sep=1.5pt] (Na) at (1,  0) {1};
            \node [draw, circle, fill=uofgpumpkin, inner sep=1.5pt] (Nb) at (1, -1.5) {2};
            \node [draw, circle, fill=uofglawn, inner sep=1.5pt] (Nc) at (0, -3) {3};
            \node [draw, circle, fill=uofgthistle, inner sep=1.5pt] (Nd) at (2, -3) {4};

@@

            \node [anchor=center, font=\Large] (A1) at (-1.5, -1.5) { $\hookleftarrow$ };
            \node [anchor=center, font=\Large] (A2) at ( 3.5, -1.5) { $\rightarrowtail$ };
        \end{scope}
        \begin{scope}[xshift=13cm]
            \foreach \n in {1, ..., 6}{
                \node [draw, circle, fill=uofgcobalt, inner sep=1pt] (N1\n) at (\n, 0) {\n};
            }

@@

    \caption{On the left, there is a non-induced isomorphism from the pattern graph, in the center, to the target
    on the right, mapping vertex 1 to 1, 2 to 3, 3 to 5 and 4 to 6. This is not an induced
    isomorphism, since there is an edge between 1 and 5 in the target but not between 1 and 3 in the
    pattern. The mapping from the pattern to the (same) target on the left, sending 1 to 2, 2 to 6, 3 to
    5 and 4 to 3, is both non-induced and induced.
    To the right, the association graph encoding for the induced version: the highlighted clique
    corresponds to the same mapping.}
    \label{figure:sip}
\end{figure}

Despite these successes, subgraph isomorphism algorithms cannot handle \emph{arbitrary} instances
involving hundreds or thousands of vertices. Experimental evaluations of subgraph isomorphism
algorithms are usually performed using a mix of real-world graphs, graphs that encode biochemistry
and computer vision problems, and randomly generated graph pairs. Using random instances to evaluate
algorithm behaviour can be beneficial, because it provides a way of generating many instances
cheaply, and reduces the risk of over-fitting when tuning design parameters. The random instances
used in each case came from common datasets \citep{DeSanto:2003,Zampelli:2010}, which were generated
by taking a random subgraph of a random (Erd\H{o}s-R\'enyi, scale-free, bounded degree, or mesh)
graph and permuting the vertices. Such instances are guaranteed to be
satisfiable---\citet{Anton:2009} exploited this property to create large sets of random satisfiable
boolean satisfiability instances.  This has been by far the most common approach used so far to
generate random subgraph isomorphism instances so far, meaning existing benchmark suites contain
relatively few non-trivial unsatisfiable instances.  Also, the satisfiable instances tend to be
computationally fairly easy, with most of the difficulty being in dealing with the size of the
model. This has lead to bias in algorithm design, to the extent that some proposed techniques will
\emph{only} work on satisfiable instances \citep{Battiti:2007}.

Our first contribution is to present and evaluate a new method for creating random pattern/target
pairs. This method generates both satisfiable and unsatisfiable instances, and can produce
computationally challenging instances with only a few tens of vertices in the pattern, and 150
vertices in the target. This is not entirely straightforward---the lack of unsatisfiable instances
for testing purposes cannot be addressed simply by using taking a pattern graph from one of the
existing random suites with the ``wrong'' target graph, as this tends to give either a trivially
unsatisfiable instance, or a satisfiable instance. (In particular, it is \emph{not} the case that a
relatively small random graph is unlikely to appear in a larger random graph.)

This work builds upon the phase transition phenomena observed in satisfiability and graph colouring
problems first described by \citet{DBLP:conf/ijcai/CheesemanKT91} and \citet{Mitchell:1992}.  For
subgraph isomorphism we identify three relevant control parameters: we can independently alter the
edge probability of the pattern graph, the edge probability of the target graph, and the relative
orders (number of vertices) of the pattern and target graphs.  For non-induced isomorphisms, with
the correct choice of parameters we see results very similar to those observed with boolean
satisfiability problems: there is a phase transition (whose location we can predict) from
satisfiable to unsatisfiable, we see a solver-independent complexity peak occur near this phase
transition, and understanding this behaviour helps us to select variable and value ordering
heuristics.

@@

When labels on vertices are introduced, as is commonly seen in graph database systems, richer
behaviour emerges. We look at some of the datasets commonly used to test graph database systems,
which the literature suggests are extremely hard to solve. Our theory predicts that these instances
should be trivial, and indeed we see that the entire perceived difficulty of every dataset we
investigate comes down to the widespread use of a particular subgraph isomorphsim algorithm which
does not use a theoretically-strong variable ordering heuristic. Our second contribution is to show
that the use of the right variable ordering heuristic, even without the more sophisticated recent
advances in subgraph algorithms (which have been largely neglected outside of the artificial
intelligence community), makes the entire graph database filter / verify paradigm unnecessary.
Finally, we explain why filter / verify cannot be beneficial even on more challenging instances
except if an extremely poor choice of subgraph isomorphism algorithm is made.

@@

Throughout, our graphs are undirected, and do not have any loops.  The \emph{order} of a graph is
the cardinality of its vertex set. We write $\operatorname{V}(G)$ for the vertex set of a graph $G$.
The \emph{complement} of a graph $G$, denoted $\shortoverline{G}$, is the graph with the same vertex
set as $G$, and with an edge between distinct vertices $v$ and $w$ if and only if $v$ and $w$ are
not adjacent in $G$. We write $G(n, p)$ for an Erd\H{o}s-R\'enyi random graph with $n$ vertices, and
an edge between each distinct pair of vertices with independent probability $p$.

A graph may have labels on its vertices, in which case we write $\ell(v)$ for the label of a given
vertex $v$. We write $G(n, p, l)$ for the graph $G(n, p)$ with vertex labels selected uniformly at
random from a set of $l$ different labels.

A \emph{non-induced subgraph isomorphism} from a graph $P$ (called the \emph{pattern}) to a graph
$T$ (called the \emph{target}) is an injective mapping from $\operatorname{V}(P)$ to
$\operatorname{V}(T)$ which preserves adjacency---that is, for every adjacent $v$ and $w$ in
$\operatorname{V}(P)$, the vertices $i(v)$ and $i(w)$ are adjacent in $T$. If vertices are labelled,
the mapping must preserve labels, so $\ell(v) = \ell(i(v))$. An \emph{induced subgraph isomorphism}
additionally preserves non-adjacency---that is, if $v$ and $w$ are not adjacent in $P$, then $i(v)$
and $i(w)$ must not be adjacent in $T$. We use the notation $i : P \rightarrowtail T$ for a
non-induced isomorphism, and $i : P \hookrightarrow T$ for an induced isomorphism. Observe that an
induced isomorphism $i : P \hookrightarrow T$ is a non-induced isomorphism $i : P \rightarrowtail T$
which is also a non-induced isomorphism $\shortoverline{i} : \shortoverline{P} \rightarrowtail
\shortoverline{T}$.

@@

\begin{figure}[tb]
    \centering\includegraphics{gen-graph-phase-transition.pdf}
    \caption{With a fixed pattern graph order of 20, a target graph order of 150, a target edge
        probability of 0.40, and varying pattern edge probability, we observe a phase transition and
        complexity peak with the Glasgow solver in the non-induced variant. Each point represents
        one instance. The lines show mean search effort and mean proportion satisfiable.}
    \label{figure:phase-transition}
\end{figure}

Suppose we arbitrarily decide upon a pattern graph order of 20, a target graph order of 150, a fixed
target edge probability of 0.40, and no vertex labels. As we vary the pattern edge probability from
0 to 1, we would expect to see a shift from entirely satisfiable instances (with no edges in the
pattern, we can always find a match) to entirely unsatisfiable instances (a maximum clique in this
order and edge probability of target graph will usually have between 9 and 12 vertices). The dark
line in \cref{figure:phase-transition} shows that this is the case. For densities of 0.67 or
greater, no instance is satisfiable; with densities of 0.44 or less, every instance is satisfiable;
and with a density of 0.55, roughly half the instances are satisfiable.

@@

\begin{figure}[tb]
    \centering\includegraphics{gen-graph-non-induced.pdf}
    \caption{Behaviour of algorithms on the non-induced variant. For each plot, the x-axis is the
        pattern edge probability and the y-axis is the target edge probability, both from 0 to 1.
        Along the top row, we show the proportion of instances which are satisfiable; the white
        bands shows the phase transitions, and the black lines are our predictions of where the
        phase transition will occur. On the final three rows, we show the number of search nodes used by the
        Glasgow, LAD and VF2 solvers; the dark regions indicate ``really hard'' instances.}
    \label{figure:non-induced}
\end{figure}

@@

\begin{figure}[p]
    \centering\includegraphics{gen-graph-induced.pdf}
    \caption{Behaviour of algorithms on the induced variant, shown in the style
    of \cref{figure:non-induced}. The fifth row plots constrainedness: the darkest region is where
    $\kappa = 1$, and the lighter regions show where the problem is either over-
    or under-constrained. The final row shows when the Glasgow algorithm performs better when given
    the complements of the pattern and target graphs as inputs---the solid lines show the location
of the phase transition, and the dotted lines are $t_d=0.5$ and the $p_d=t_d$ diagonal.}\label{figure:induced}
\end{figure}

\begin{figure}[tb]
    \centering\includegraphics{gen-graph-sat.pdf}
    \caption{Behaviour of other solvers on the induced variant on smaller graphs, shown in the style of
        \cref{figure:non-induced}. The second row shows the number of search nodes used by the
    Glasgow algorithm, the third and fourth rows show the number of decisions made by the pseudo-boolean
and SAT solvers, and the final shows the number of search nodes used on the clique
encoding.}\label{figure:alt}
\end{figure}

@@

For larger patterns, more complex behaviour emerges. If we are in the intersection of the bottom half
and the bottom right diagonal of the search space, we should always retain the original heuristic,
and if we are in the intersection of the top half and the top left diagonal, we should always use
the complements. This behaviour can be predicted by taking the partial derivatives of $\langle Sol
\rangle$ in the $-p_d$ and $t_d$ directions.  However, when inside the remaining two eighths of the
parameter space, the partial derivatives of $\langle Sol \rangle$ disagree on which heuristic to
use, and using directional derivatives is not enough to resolve the problem. A close observation of
the data suggests that the actual location of the phase transition may be involved (and perhaps
\citeS{Walsh:1998} suggestion applies only in these conditions). In any case, $\langle Sol
\rangle$ is insufficient to explain the observed behaviour in these two eighths of the parameter space.

@@

\section{Other Models of Randomness}\label{section:models}

\section{Labelled Graphs}\label{section:labelled}

\begin{figure}[t]
    \centering\includegraphics{gen-graph-labels.pdf}
    \caption{On the top row, predicted and actual location of the phase transition for labelled
    non-induced random subgraph isomorphism, with a pattern order of 20, a target order of 100,
    varying pattern (x-axis) and target (y-axis) density, and varying numbers of labels. On the second
    row, the average number of search nodes needed to solve an instance.}\label{figure:labels}
\end{figure}

So far, we have looked at unlabelled graphs. What happens when labels on vertices are introduced?
This is common in real-world applications---for example, when working with graphs representing
chemical molecules, mappings are typically expected only to match carbon atoms with carbon atoms,
hydrogen atoms with hydrogen atoms, and so on.  We will look at the non-induced variant, as this
seems to be more common in the literature, and return to the Erd\H{o}s-R\'enyi model.

Suppose our labels are drawn randomly from a set $L = \{ 1, \ldots, k \}$, where $k$ is reasonably
small compared to $p$. We can partition the pattern vertices by label into disjoint sets \[\{
    \operatorname{V}(P)\vert_1 = \{ v \in \operatorname{V}(P) : \ell(v) = 1 \}, \ldots,
    \operatorname{V}(P)\vert_k \}\textnormal{\,,}\] each of which is expected to contain
$\nicefrac{p}{k}$ vertices. Similarly, we may partition the target vertices into disjoint sets $\{
    \operatorname{V}(T)\vert_1, \ldots, \operatorname{V}(T)\vert_k \}$.

Without labels, there are $t^{\underline{p}} = t \cdot (t - 1) \cdot \ldots \cdot (t - p + 1)$
possible injective assignments of target vertices to pattern vertices.  With labels, observe that
for any label $x$, vertices in $\operatorname{V}(P)|_x$ may only be mapped to vertices in
$\operatorname{V}(T)|_x$.  Thus for each label $x$, we have an expected $\nicefrac{p}{k}$ variables,
each of whose domains contain $\nicefrac{t}{k}$ values. We would like to say that the size of the
state space is now \[ |S| = \left((\nicefrac{t}{k})^{\underline{\nicefrac{p}{k}}}\right)^{k}
\textnormal{,} \] but to do this we must state what $a^{\underline{b}}$ means when $b$ is fractional.
A reasonable continuous extension may be obtained by taking \[ |S| = \left(\frac{\Gamma\left(\nicefrac{t}{k}
+ 1\right)}{\Gamma\left(\nicefrac{t}{k} - \nicefrac{p}{k} + 1\right)}\right)^{k} \textnormal{\,.} \]

As before, we expect the pattern to have $d_p \cdot \binom{p}{2}$ edges, so we obtain the
probability of all of these edges being mapped to edges in the target by raising $d_t$ to this
power, giving an expected number of solutions of \[ \langle Sol \rangle = \left(
    \frac{\Gamma\left(\nicefrac{t}{k} + 1\right)}{\Gamma\left(\nicefrac{t}{k} - \nicefrac{p}{k} +
1\right)} \right)^{k}  \cdot
{d_t}^{d_p \cdot \binom{p}{2}} \textnormal{\,.} \]

So how good are these predictions? We draw some heatmaps in \cref{figure:labels}. For small numbers
of labels, our predictions are slightly better than in the unlabelled case: there seems to be less
of a variance problem for very dense patterns. Even as the numbers of labels becomes relatively
large, the prediction of the phase transition still occurs in the right place, but we start to see
sporadic unsatisfiable instances deep inside the satisfiable reason. With sufficiently many labels,
we get a kind of phase transition from ``all unsatisfiable'' to ``mixed satisfiable and
unsatisfiable''.

\subsection{Flawed Instances}

What goes on here?

Min unsat core.

Insufficient vertices.

Two adjacent differently-coloured vertices.

VF2 does not detect the case that there are insufficient, or even no, labels.

\section{Querying Graph Databases}\label{section:filterverify}

The general problem is, for a set of target graphs, to process a pattern query and return every
target graph which is subgraph-isomorphic to that pattern. The set of target graphs is usually seen
as fixed, or at least rarely-changing, whilst the patterns arrive dynamically. This has lead to the
development of systems which perform extensive computations on the target graphs beforehand, in the
hopes of reducing the response times for individual pattern queries. The most popular of these
strategies is a form of indexing which is often named \emph{filter / verify}.

\subsection{The Filter / Verify Paradigm}

The filter / verify approach has an interesting history, of which we now give a very selective and
incomplete overview. Our description is biased by a general modern understanding of \NP-completeness
which was not widely known at least at the time of the earlier papers we discuss. The common theme
of all of the following papers is that pre-computed information is used to eliminate certain
unsatisfiable instances from consideration, without performing a subgraph isomorphism test.

An early graph database system by \citet{DBLP:conf/pods/ShashaWG02} uses a filtering heuristic to
eliminate unsatisfiable instances based upon simple structural elements. It is not clear whether the
aim is to reduce I/O costs, or to reduce the number of queries which must be tested, and the
experiments do not answer whether the indexing is effective. (?? This isn't the first paper on this,
but before this seem to be commercial applications, in particular a system called Daylight. Maybe
more digging is required.)

Subsequently, in a widely cited paper, \citet{DBLP:conf/sigmod/YanYH04} introduce an indexing system
called gIndex. Again, this system handles queries by first producing a set of candidates by
eliminating certain unsatisfiable instances, this time by using substructures. They argue that the
query response time, which is to be minimised, is governed by the equation \[ T =
T_{\mathit{search}} + \left|C_q\right| \cdot T_{\mathit{iso\_test}} \textnormal{,}\] where
$T_{\mathit{search}}$ is the time taken to search for a candidate set of potential solutions of size
$\left|C_q\right|$, and $T_{\mathit{iso\_test}}$ is the cost of a subgraph isomorphism test. They
reason that since isomorphism testing is \NP-complete, by making $\left|C_q\right|$ as small as
possible, the query response time will be reduced.  (Importantly, it is \emph{not} time taken to
load graphs from disk which contributes to the per-candidate cost, but rather the time to perform a
subgraph isomorphism call.) They conclude that ``graph indexing plays a critical role at efficient
query processing in graph databases''.

This equation is repeated and expanded upon by \citet{DBLP:journals/tods/YanYH05} to explicitly
separate I/O and isomorphism testing costs, obtaining a query response time of \[
    T_{\mathit{search}} + \left|C_q\right| \cdot (T_{\mathit{io}} + T_{\mathit{iso\_test}})
    \textnormal{.} \] The authors explicitly state that ``the value of $T_{\mathit{iso\_test}}$ does
not change much for a given query''. The argument presented is that
\begin{displayquote}``Sequential scan is very costly
because one has to not only access the whole graph database but also check subgraph isomorphism. It
is known that subgraph isomorphism is an \NP-complete problem. Clearly, it is necessary to build
graph indices in order to help processing graph queries.''\end{displayquote}

With what we now know about the behaviour of modern subgraph isomorphism algorithms, and the nature
of solving \NP-complete problems in general, we should immediately be suspicious of these claims. We
do not expect $T_{\mathit{iso\_test}}$ to be anything like a constant, even if the orders of the
input graphs are similar. In particular, any instance which can be excluded based upon filtering
must have a very small proof of unsatisfiability. These instances \emph{should} be trivial with any
decent matching algorithm. Thus, all filtering \emph{should} be doing is eliminating the startup
costs of a trivial subgraph isomorphism call. The fact that filtering was successful empirically
should make us wonder whether the subgraph isomorphism algorithm being used was excessively
primitive. Indeed, the matching algorithm used is described only as ``the simplest approach'' in the
paper.  Additionally, the experiments focus on reducing the size of the candidate set, without
considering the time taken to verify different candidate set instances.  The claim that
$T_{\mathit{iso\_test}}$ does not change much is not justified experimentally, and no consideration
is given as to whether this would hold true for other subgraph matching algorithms.

Moving forwards, the (simpler form of the) query response time equation is repeated by
\citet{DBLP:conf/vldb/ZhaoYY07}. Again, the work has a focus on reducing the candidate set size
through indexing. The authors appear to believe that the cost of the isomorphism test is not a major
factor in influencing the result, and focus on the remaining terms in the equation. They use the
``average cost'' of a subgraph isomorphism test as a constant, without considering that the average
cost could be influenced by the character of the candidate set.

The equation is also used by \citet{DBLP:conf/icde/JiangWYZ07}, who claim that ``usually the
verification time dominates the Query Response Time [s]ince the computational complexity of
$T_{\mathit{iso\_test}}$ is NP-Complete''. They note that \begin{displayquote}``Approximately, the
    value of $T_{\mathit{iso\_test}}$ does not change too much with the difference of query. Thus,
the key to reducing query response time is to minimize the size of the candidate answer
set''.\end{displayquote} This claim becomes understandable when one examines the subgraph matching
algorithm chosen for the verification step \citep{DBLP:journals/jacm/Ullmann76}: as the algorithm
predates techniques like arc-consistent all-different propagation \citep{DBLP:conf/aaai/Regin94}, it
cannot immediately detect unsatisfiability in simple cases like there being two red vertices in the
pattern but only one in the target.

Without using the equation, \citet{DBLP:conf/sigmod/ChengKNL07} state that since subgraph
isomorphism is NP-complete, processing by a sequential scan is infeasible. They introduce new
filtering techniques to try to avoid the subgraph isomorphism step. A similar claim is made by
\citet{DBLP:conf/icde/ZhangHY07} in a introduction of another indexing technique: ``obviously it is
inefficient to perform a sequential scan on every graph in the database, because the subgraph
isomorphism test is expensive''.

Muddying the waters slightly, a survey by \citet{DBLP:journals/datamine/HanCXY07} states that
``large volumes of data'' (not \NP-completeness) is the reason for using indexing in these systems.
However, in a description of a system tailored to biological networks,
\citet{DBLP:conf/edbt/ZhangLY09} state that \begin{displayquote}``Since the size of the raw database
    graph is small, it can be easily fit in the main memory.  However, the query (matching) time
will be very long due to the NP-hard complexity.''\end{displayquote} They suggest that indexing is
a way of avoiding this.

\citet{DBLP:journals/tkde/WangWYY12} look at indexing large sparse graphs. They state that ``because
subgraph isomorphism is an \NP-complete problem, a filter-and-verification method is usually
employed to speed up the search efficiency of graph similarity matching over a graph set''.
Similarly, after reviewing the literature, \citet{DBLP:journals/vldb/YuanM13} argue that subgraph
querying is costly because it is \NP-complete, and that indices can improve the performance of graph
database queries. Again, new indexing techniques are introduced.  Subsequently,
\citet{DBLP:journals/pvldb/KatsarouNT15} perform a comprehensive comparison of graph database
filtering techniques. They state that performing a query against each graph in the dataset
``obviously does not scale, as subgraph isomorphism is \NP-complete''.

The filter / verify paradigm also influences other research. For example, recently
\citet{DBLP:journals/pvldb/YuanMG13} continue a line of supergraph search work, again using a
filtering step to avoid subgraph isomorphism calls. There is also research into maintaining indices
when the set of target graph changes: for example \citet{DBLP:conf/sigmod/YuanMYG15} look at
algorithms for updating graph indices. And in describing a system for reusing results of queries
which are sub- or super-graphs of previous queries, \citet{DBLP:conf/edbt/WangNT16} state that
querying is a ``very costly operation as it entails the \NP-complete problem of subgraph
isomorphism'', and place ``an emphasis on the number of unnecessary subgraph isomorphism tests''.
